---
title: "Basic_popgen_stats"
author: "Dylan Cohen"
date: "5/6/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r library}
library(poppr)
library(hierfstat)
library(gaston)
library(vcfR)
library(reshape2)
library(pegas)
library(cowplot)
library(adegenet)
library(plotly)
library(directlabels)
library(ggthemes)
library(dartR)
```

#Set working directory to where the vcf files are found

```{r}
setwd("/Users/dcohen/Desktop/R_Stuff")
```


#Here we read in the  VCF files and convert them to genlight object
```{r read in vcf with vcfR}
vcf<-read.vcfR("Tharpii_only.vcf")

vcf_fug<-read.vcfR("Fugatei_only.vcf") 
vcf_long<-read.vcfR("Longiflora_only.vcf") 
vcf_all<-read.vcfR("All_species.vcf") 

tharp_gl <- vcfR2genlight(vcf)
fug_gl<-vcfR2genind(vcf_fug)
long_gl<-vcfR2genind(vcf_long)
All_gl<-vcfR2genind(vcf_all)
```

#Assign each individual to a population
```{r -create popmap}
pop(tharp_gl)<- as.factor(c("BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	
                             "BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",
                             "BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",
                             "BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN","CAP","CAP","CAP","CAP",	"CAP","CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP","CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",
                             "CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",
                             "CAP",	"CAP",	"CAP","CPC",	"CPC",
                            "CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",
                            "CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",
                            "CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",
                            "CPC",	"CPC",	"CPC",	"RED",	"RED",	"RED",	"RED",	"RED",
                             "RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",
                             "RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",
                             "RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",
                             "RED",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",
                             "Tex",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",
                             "Tex",	"Tex",	"Tex",	"Tex",	"Tex"))

```

```{r - VCF files for just fugatei}
pop(fug_gl)<- as.factor(c("BOS" ,"BOS", "BOS", "BOS" ,"BOS", "BOS", "BOS",
                             "BOS" ,"BOS", "BOS", "FIF", "FIF", "FIF" ,"FIF",
                             "FIF", "FIF" ,"FIF", "FIF", "FIF", "FIF" ,"SEV", 
                             "SEV", "SEV", "SEV" , "SEV",  "SEV" , "SEV" , "SEV", 
                             "SEV" , "SEV"))
```

```{r - VCF file for just longiflora}
pop(long_gl)<-as.factor(c("AZO", "AZO", "AZO", "AZO", "AZO", "AZO" ,"AZO" ,"AZO", "AZO", "AZO",
"BLA","BLA","BLA","BLA","BLA","BLA","BLA","BLA","DEL","DEL","DEL", "DEL", "DEL", "DEL", "DEL",  "DEL","DEL","DEL","ENG","ENG",  "ENG","ENG","ENG","ENG","ENG","ENG","ENG","ENG","OVI","OVI","OVI","OVI","OVI","OVI","OVI" ,"OVI","OVI","OVI","OVI","OVI","WOO","WOO","WOO","WOO" ,"WOO","WOO","WOO", "WOO"))
```


```{r - this is for VCF files with all 3 species}
pop(All_gl)<- as.factor(c("Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp", "Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp", "Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp", "Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp",	"Tharp","Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei","Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",
"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei",	"Fugatei","Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora","Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora","Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora","Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",
"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora","Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora",	"Longiflora","Longiflora"))
```
                             
```{r- compliance check to make sure file is useable for dartr}
tharp_gl2 <- gl.compliance.check(tharp_gl)
fug_gl2 <- gl.compliance.check(fug_gl)
long_gl2 <- gl.compliance.check(long_gl)
All_gl2 <- gl.compliance.check(All_gl)
```

                             
```{r - generate observed heterzygosity per individual}
gl.report.heterozygosity(tharp_gl2,method = 'ind')
gl.report.heterozygosity(fug_gl2,method = 'ind')
gl.report.heterozygosity(long_gl2,method = 'ind')
gl.report.heterozygosity(All_gl2 ,method = 'ind')
```

#To build box plots copy and paste the output or save it to a CSV file in R, we only need the first column (names) and second column (observed heterzygozity) below is an example with just tharpii output

```{r-Tharpii only boxplot}
data3<-read.csv("tharp_pops.csv")
data3$X <- factor(data3$X , levels=c("BEN", "RED", "CPC", "CAP","TEX"))
boxplot(data3$exp~data3$X, data3 = data,  ylab="Expected He", xlab = "Population",cex.axis=1.5, cex.lab=1.5)
```

#To get inbreeding coefficient for each population (Tharpii only below) using hierfstat
```{r-convert the vcf into genid object}

tharp_gen<-vcfR2genind(vcf)

#assign each individual to a population
pop(tharp_gen)<- as.factor(c("BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	
                             "BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",
                             "BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",
                             "BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"BEN",	"CPC",	"CPC",
                             "CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",
                             "CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",
                             "CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",	"CPC",
                             "CPC",	"CPC",	"CPC",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",
                             "CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",
                             "CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",
                             "CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",	"CAP",
                             "CAP",	"CAP",	"CAP",	"RED",	"RED",	"RED",	"RED",	"RED",
                             "RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",
                             "RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",
                             "RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",	"RED",
                             "RED",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",
                             "Tex",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",	"Tex",
                             "Tex",	"Tex",	"Tex",	"Tex",	"Tex"))	

#Convert the genid object to a hierfstat object
tharp_hfstat<-genind2hierfstat(tharp_gen)
#Run basic stats using  hierfstat
basicstat<-basic.stats(tharp_hfstat, diploid = TRUE, digits = 3)
#move results into a data frame
overall<-as.data.frame(basicstat$overall)
#get the mean for each population
Tharpii_Fis <-colMeans(basicstat$Fis,na.rm = T)
#Displays mean Fis for each pop
Tharpii_Fis 
```

#Next lets get pairwise Fst for all populations, we already have all three species as genlight dataframes and so all we need to do is plug them into the dartr function gl.fst
```{r-pairwise Fst}
# tharpii
tharp_fst <- gl.fst.pop(tharp_gl)
D_tharp <- as.dist(tharp_fst$Fsts, diag = FALSE, upper = FALSE)
D_tharp

# fugatei
tf <- gl.fst.pop(fug_gl)
D_fug <- as.dist(tf$Fsts, diag = FALSE, upper = FALSE)
D_fug

long_fst <- gl.fst.pop(long_gl)
D_long <- as.dist(long_fst$Fsts, diag = FALSE, upper = FALSE)
D_long 
```
#Isolation by distance plots - because fugatei and tharpii are sensitive species and I dont want to publicly post their locality information I will only do this for the common species longiflora

```{r-isolation by distance plot}
library(geodist)
# load the population coordinates for longiflora
list3 <- data.frame(longitude = c(-104.546147,
                                  -105.48917,
                                  -105.16436,
                                  -107.144,
                                  -105.912392,
                                  -105.15552), 
                    latitude = c(32.390842, 
                                 32.04322,  
                                 32.18425,
                                 33.13,
                                 32.747678,
                                 32.54203),
                    name= c('AZO', 
                            'BLA', 
                            'DEL',
                            'ENG',
                            'OVI',
                            'WOO'))

#Create a pairwise distance matrix
distance_matrix3 <- geodist(list3, measure = 'geodesic' )/1000 
colnames(distance_matrix3) <- list3$name
rownames(distance_matrix3) <- list3$name
Dgeo3 <- distance_matrix3
Dgeo3 

#plot Fst x distance and draw line of best fit
plot(Dgeo3[lower.tri(Dgeo3)], D_long, xlab = "Euclidean distance (km)", ylab = "Population Pairwise Fst",  pch=20, cex=3, col="black", ylim =c(0, 0.15), xlim =c(0,275),cex.axis=1.5, cex.lab=1.5)
abline(lm(D_long ~ Dgeo3[lower.tri(Dgeo3)]),col="purple", lwd =3)
```



                             
                             

